---
title: "IRMAA and Tax Cliffs"
author: "dchilders"
format: html
editor: source
editor_options: 
  chunk_output_type: console
---

Many retirees have a misconception that Medicare coverage is free. While Medicare A (insurance for hospital stays) is free for most people, there are monthly premiums for Medicare B (outpatient care) and Medicare Part D (prescription drugs).

Let's look at a table for the common monthly premiums for the different medicare coverages:

Created at `r Sys.time()`

```{r setup}
#| include: false
library(tidyverse)
library(gt)
library(emoji)
library(scales)
library(irslimits) # grab data from 'irslimits' package
```

some text here

```{r premium}
doc_emoji <- emoji::emojis %>%
  unnest_longer(keywords) %>%
  filter(name == "woman health worker") %>%
  slice(1) %>%
  pull(emoji)

d <- tribble(
  ~part, ~coverage, ~monthly_premium,
  "A", emoji("hospital"), 0,
  "B", doc_emoji, 185,
  "D", emoji("pill"), 30
)

#gt(d) %>% fmt_currency(monthly_premium, decimals = 0)
```

In 2025, \$185 is the lowest monthly premium for Part B Medicare Coverage. Individuals and couples with high incomes have to pay an income-related monthly adjustment amount (IRMAA).

```{r part_b_define}
# part_b <- tribble(
#   ~year, ~premium, ~deductible,
#   2024L, 174.70, 240,
#   2025L, 185.00, 257
# )

data(premiums_b)


```

```{r irmaa_table}
irmaa %>%
  filter(filing == "single") %>%
  mutate(
    magi = paste(scales::dollar(magi_low), scales::dollar(magi_high), sep = " - "),
    magi = ifelse(magi == "$500,001 - Inf", "$500,001+", magi)
    ) %>%
  select(magi, adjustment, total_premium) %>%
  gt() %>%
  tab_header(
    "IRMAA Adjustment by Income",
    subtitle = "For single filer in 2025"
    ) %>%
  fmt_currency(c(adjustment, total_premium)) %>%
  cols_label(
    magi = "Income",
    adjustment = md("IRMAA<br>Adjustment"),
    total_premium = md("Part B<br>Monthly Premium")
  )
```

Approximately 8% of Medicare Part B enrollees have incomes high enough to be [impacted](https://www.cms.gov/newsroom/fact-sheets/2025-medicare-parts-b-premiums-and-deductibles?utm_source=chatgpt.com) by IRMAA.


A single medicare beneficiary with an income of \$106,000 would pay \$185 x 12 = \$2,220 in Medicare Part B Premiums per year. If her income increases by just \$1 to \$106,001 per year, then the first IRMAA threshold kicks in, and she would pay \$259 x 12 = \$3,108.

That one extra dollar of income caused her to pay \$888 in extra medicare premiums. When filing taxes, we're used to marginal income tax rates of 10%, 12%, 22%, 24%, 32%, 35%, and 37%. For the purposes of medicare premiums, that last dollar of income has a marginal tax rate of 88,800%?

This is known as a tax cliff, where you pay a disproportionately high amount of tax for an extra dollar of income.

To visualize the cliff, let's look at a graph of how the annual IRMAA premiums vary by income.

```{r}
single <- irmaa %>% filter(filing == "single")

ggplot(single, aes(x = magi_low, xend = magi_high, y = total_premium * 12)) +
  geom_segment() +
  theme_minimal() +
  expand_limits(y = 0, x = 600000) +
  scale_y_continuous(labels = label_currency(suffix = "K", scale = .001), breaks = pretty_breaks(7)) +
  scale_x_continuous(labels = label_currency(suffix = "K", scale = .001), breaks = pretty_breaks(9)) +
  geom_point(aes(x = magi_low, y = total_premium * 12)) +
  labs(
    title = "How does income impact Medicare Part B premiums?",
    #subtitle = "By Trailing 2 year MAGI (modified adjusted gross income)",
    x = "MAGI (Income Two Years Prior)",
    y = "Premium"
  ) +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5, hjust = 0))
```

The five vertical jumps in the plots are the IRMAA cutoffs. 

<!--
Instead of the annual premium in dollars, we could look at the premium *rate*, the percentage of income that is paid in medicare premiums.
-->

```{r create_premium_ratedata}
#| include: false
dd <- tibble(magi = 0:7e5)

dd2 <- dd %>%
  left_join(
    single %>% select(magi_low, total_premium),
    by = c("magi"="magi_low")
  )

dd3 <- dd2 %>%
  fill(total_premium) %>%
  mutate(
    annual_premium = total_premium * 12,
    premium_rate = annual_premium / magi
    ) 

dd4 <- dd3 %>%
  mutate(
    medicare_amt = (0.0145 * magi) + (0.009 * pmax(0, magi - 250000)),
    medicare_rate = medicare_amt / magi
  )

dd5 <- dd4 %>%
  select(magi, premium_rate, medicare_rate) %>%
  pivot_longer(-magi, names_to = "rate_type", values_to = "rate") %>%
  mutate(rate_type = str_remove_all(rate_type, "_rate")) %>%
  filter(magi >= 70000) %>% print

#dd5 %>% filter(rate_type == "premium", between(magi, 105995, 106005))

dd5 %>%
  filter(rate_type == "premium") %>%
  #filter(magi >= 100000) %>%
  ggplot(aes(x = magi, y = rate)) +
  geom_line(data = dd5 %>% filter(rate_type == "premium", magi <= 106000)) +
  geom_line(data = dd5 %>% filter(rate_type == "premium", between(magi, 106001, 133000))) +
  geom_line(data = dd5 %>% filter(rate_type == "premium", between(magi, 133001, 167000))) +
  geom_line(data = dd5 %>% filter(rate_type == "premium", between(magi, 167001, 200000))) +
  geom_line(data = dd5 %>% filter(rate_type == "premium", between(magi, 200001, 500000))) +
  geom_line(data = dd5 %>% filter(rate_type == "premium", magi >= 501000)) +
  theme_minimal() +
  expand_limits(y = 0, x = 700000) +
  scale_y_continuous(labels = label_percent(), breaks = pretty_breaks(7)) +
  scale_x_continuous(labels = label_currency(suffix = "K", scale = .001), breaks = pretty_breaks(9)) +
  #geom_point(aes(x = magi_low, y = total_premium * 12)) +
  labs(
    title = "What are premiums as a percentage of MAGI?",
    #subtitle = "By Trailing 2 year MAGI (modified adjusted gross income)",
    x = "MAGI (Income Two Years Prior)",
    y = "Premium\nRate"
  ) +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5))
```

Most of the tax code doesn't work this way. Here are the 2025 federal income tax rates for single filers:

```{r single_tax_rates}
single_income_rates <- tribble(
  ~income_low, ~income_high, ~rate, 
  0, 11925, 0.10,
  11926, 48475, 0.12,
  48476, 103350, 0.22,
  103351, 197300, 0.24,
  197301, 250525, 0.32,
  250526, 626350, 0.35,
  626351, Inf, 0.37
) %>% print

gt(single_income_rates)

ii <- tibble(income = 0:7e5)

# show plot of income tax rates
ii2 <- ii %>%
  mutate(
    income_tax = 0.1 * pmin(income, 11925) + 0.12 * pmin(income - 11925, 48476)
  ) %>% 
  print
    
    dd2 <- dd %>%
  left_join(
    single %>% select(magi_low, total_premium),
    by = c("magi"="magi_low")
  )

dd3 <- dd2 %>%
  fill(total_premium) %>%
  mutate(
    annual_premium = total_premium * 12,
    premium_rate = annual_premium / magi
    ) 

```



<!-- So most beneficiaries pay \$185 x 12 = \$2,220 per year and *very* high earners pay \$628.90 x 12 = #\$7,546.8 per year for Medicare Part B. -->


Most beneficiaries

# Define a tax cliff

```{r}

```

```{r}

```
